Interacting with Modal Logics \\ in the Coq Proof Assistant

### 2014-12-12 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Add missing references

\author{
  Christoph Benzm\"{u}ller\inst{1}\thanks{Supported by the German Research Foundation (DFG) under grant BE 2501/9-1.} 
  \and 
  Bruno Woltzenlogel Paleo\inst{2}
}

@@

\begin{abstract}  
  This paper describes an embedding of higher-order modal logics in
  the \Coq proof assistant. {\Coq}'s capabilities are extended in a
  minimalistic manner, which is nevertheless sufficient for the
  formalization of significant, non-trivial modal logic proofs. The
  elegance, flexibility and convenience of this approach, from a user
  perspective, are illustrated here with the successful formalization
  of G\"odel's ontological argument.
\end{abstract}

@@

Another recently explored possibility is the embedding of first-order
and even higher-order modal logics into classical higher-order logics
\cite{J23,B9,J22}, for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist. The embedding approach is
flexible, because various modal logics (even with multiple modalities
or varying/cumulative domain quantifiers) can be easily supported by
stating their characteristic axioms. Moreover, the approach is
relatively simple to implement, because it does not require any
modification in the source code of the higher-order prover. The prover
can be used as is, and only the input files provided to the prover
must be especially encoded. Furthermore, the efficacy and efficiency
of the embedding approach has been confirmed in philosophically
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

However, one may wonder whether the embedding approach is adequate
also for \emph{interactive} reasoning, when the user proves theorems
by interacting with a proof assistant such as \Coq\footnote{The \Coq
proof assistant was chosen merely because of the authors' greater
familiarity with the tactic language of this system. Nevertheless, the
techniques presented and discussed here are likely to be useful for
other proof assistants, such as \texttt{Isabelle} \cite{Isabelle} or
\texttt{HOL Light} \cite{HOLLight}.}. The main goal and novelty
of this paper is to study this question. Our answer is positive.

@@

In order to illustrate the use of the implemented embedding, we show
here the formalization of Scott's version \cite{ScottNotes} of G\"odel's
ontological argument for God's existence (in Section \ref{sec:proof}).
This proof was chosen mainly for two reasons. Firstly, it requires not
only modal operators, but also higher-order quantification. Therefore,
it is beyond the reach of specialized propositional and first-order
(modal) theorem provers. Secondly, this argument addresses an ancient
problem in Philosophy and Metaphysics, which has nevertheless received
a lot of attention in the last 15 years, because of the discovery of
the modal collapse \cite{sobel,sobel2004logic}. This proof lies in the center of a
vast and largely unexplored application domain for automated and
interactive theorem provers.

The simpler ontological argument of Anselm has been automatically
verified with \texttt{PVS} by Rushby \cite{Rushby} and with 
first-order theorem provers by Oppenheimer and Zalta \cite{Zalta}. G\"odel's
argument was automatically verified in our previous work on embedding-
based fully automated modal theorem proving \cite{arXiv,AFP}. This
paper presents the first fully interactive and detailed formalization
of this proof in a proof assistant. Hence we here reveal the proof
structure on a cognitively adequate level of
detail. Such detailed proof structure has been hidden in our
other papers on the subject due to the use of automated theorem
provers. In comparison to \cite{Rushby,Zalta}, however, our
contribution stands out with its surprising technical 
simplicity and elegance.

In addition to philosophy, propositional and quantified modal logics 
have (potential) applications in various other fields as well, including, for instance, 
verification, artificial intelligence
agent technologies, law and linguistics (cf.~\cite{ModalLogic} and the references therein).
Therefore, the main contribution described in this paper -- convenient techniques for leveraging a powerful proof assistant such as \Coq for interactive reasoning for modal logics -- may serve as a starting point for
many interesting projects.

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition mnot (p: o)(w: i) := ~ (p w).
\Notation "m~  p" := (mnot p) (at level 74, right associativity).

@@

\Definition mequiv (p q:o)(w:i) := (p w) <-> (q w).
\Notation "p m<-> q" := (mequiv p q) (at level 99, right associativity).

\Definition mequal (x y: u)(w: i) := x = y.
\Notation "x m= y" := (mequal x y) (at level 99, right associativity).
\end{Verbatim}

@@

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction (ND) calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions (CIC) \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper. This calculus is classical, because of the double negation elimination rule. Although CIC is intuitionistic, it can be made classical by importing \Coq's classical library, which adds the axiom of the \emph{excluded middle} and the double negation elimination lemma.}
(e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

\begin{calculus}
{Rules of a (classical) ND calculus}
{fig:NaturalDeduction}
\begin{small}

%\vspace{1em}

@@

@@

Rules for $\alpha\beta\eta$-equality and axioms (or rules) for extensionality are omitted here since they are not important for the rest of the paper. For a full, sound and Henkin-complete, classical higher-order ND calculus see \cite{HOLND}.
\end{small}

%\vspace{1em}
\end{calculus}

@@

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they give rise to the inference
rules\footnote{The ND calculus with the rules from
Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
complete relatively to the calculus of Figure
\ref{fig:NaturalDeduction} extended with a necessitation rule and the
modal axiom K \cite{ToDo:UnpublishedPaperInGithub}. Starting from a sound and Henkin-complete ND calculus for classical higher-order logic (cf. Figure~\ref{fig:NaturalDeduction}), the additional
modal rules in Figure \ref{fig:NDK} make it sound and Henkin-complete
for modal higher-order logic \textbf{K}.} shown in Figure
\ref{fig:NDK}. The labels that name boxes in those inference rules are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\s\s
\infer[\nec_I]{\nec A}{\omega: \fbox{\infer*{A}{}} }
\s
\infer[\nec_E]{w: \fbox{ \infer*{}{A} } }{\nec A}
%\vspace{2em}
\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

@@

\noindent 
The proof of this lemma is displayed as a ND
proof in  Figure \ref{fig:mp_dia}. As expected, \Coq's basic tactics
(e.g. \texttt{intros} and \texttt{apply}) work without modification.
The \texttt{intros p q H1 H2} tactic application corresponds to the
universal quantifier  and implication introduction inferences in the
bottom of the proof.  The \texttt{apply H3} tactic application
corresponds to the implication  elimination inference. The $\pos_E$,
$\pos_I$ and $\nec_E$ inferences correspond, respectively, to the
\texttt{dia\_e H1},  \texttt{dia\_i w0} and \texttt{box\_e H2 H3}
tactic applications.  The internal box named $w_0$ is accessed by
exactly one strong  modal inference, namely $\pos_E$.

\begin{figure}[H]
\centering
\includegraphics[width=0.5\textwidth]{ND-(mp_dia).png}
\caption{ND proof of \texttt{mp\_dia}\label{fig:mp_dia}}
\end{figure}

\noindent
The same lemma could be proved without the new modal tactics, as shown below.
But this is clearly disadvantageous, for several reasons: 
the proof script becomes longer;
the definitions of modal operators must be unfolded, 
either explicitly (as done below) or implicitly in the user's mind; 
tactic applications dealing with modal operators cannot be easily distinguished
from tactic applications dealing with quantifiers; and  
hypotheses about the reachability of worlds (e.g. \texttt{R1} below) 
must be handled explicitly. In summary, without the modal tactics, 
a convenient and intuitive correspondence between proof scripts and 
modal ND proofs would be missing.

@@

In order to demonstrate the efficacy and convenience of the modal
embedding approach not only for proving simple lemmas and theorems,
but also for larger developments, we include here a full and detailed
formalization of G\"odel's ontological argument.

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology. G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern
culmination of this tradition, following particularly the footsteps of
Leibniz. Various slightly different versions of axioms and definitions
have been considered by G\"{o}del and by several philosophers who
commented on his proof (cf.
\cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).
The formalization shown in this Section aims at being as similar as
possible to Dana Scott's version of the proof \cite{ScottNotes}. The
formulation and numbering of axioms, definitions and theorems is the
same as in Scott's notes. Even the \Coq proof scripts follow precisely
all the steps in Scott's notes. Scott's assertions are emphasized
below with comments. In contrast to the formalization in \Isabelle
\cite{AFP}, where automation via \texttt{Metis}
\cite{Hurd03first-orderproof} and \texttt{Sledgehammer}
\cite{Sledgehammer} using tools such as \texttt{Nitpick}
\cite{Nitpick}, \texttt{LEO-II} \cite{LEO-II} and \texttt{Satallax}
\cite{Satallax} has been successfully employed, the formalization in
\Coq used no automation. This was a deliberate choice, mainly because
it allowed a qualitative evaluation of the convenience of the
embedding approach for \emph{interactive} theorem proving. Moreover,
in order to formalize exactly Scott's version and not some arbitrary
version found automatically\footnote{The proofs found automatically
  by the above provers indeed differ from the one presented here:
  e.g., the strong S5 principle used below (and by Scott) is not needed; the
  ATP proofs only rely on axiom B.}, automation would have to be
heavily limited anyway. Furthermore, the deliberate preference for
simple tactics (mostly \emph{intro}, \emph{apply} and the modal
tactics described in Section \ref{sec:Tactics}) results in proof
scripts that closely correspond to common ND
proofs. This hopefully makes the formalization more accessible to
those who are not experts in \Coq's tactics but are nevertheless
interested in G\"odel's proof.

G\"odel's proof requires \Coq's classical logic libraries as well as the
\texttt{Modal} library developed by us and described in Sections
\ref{sec:Embedding} and \ref{sec:Tactics}.

@@

\noindent
G\"odel's theory has a single higher-order constant, \texttt{Positive}, 
which ought to hold for properties considered \emph{positive} in a moral sense.

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\com{(* Axiom A1 (divided into two directions):} 
\com{   either a property or its negation is positive, but not both *)}
\Axiom axiom1a : 
  [ mforall p, (Positive (\fun x: u => m~(p x))) m-> (m~ (Positive p)) ].

@@

Although fully automated (as opposed to interactive) theorem proving
is beyond the scope of this paper, it is worth mentioning that all
lemmas and theorems in Sections \ref{sec:Embedding},
\ref{sec:SimpleExamples} and \ref{sec:BeyondK} (but not \ref{sec:proof}) could have been proven
automatically using \Coq's \texttt{firstorder} tactic. The
implementation of hints to allow \Coq's automatic tactics to take full
advantage of the embedding and the modal axioms still remains for
future work.

The theological implications of the verified correctness of G\"odel's
proof certainly depend on a critical discussion of the underlying
concepts, definitions and axioms, which is beyond the scope of this
paper.  Clearly, the application of theorem proving technology to the
domain of theoretical philosophy can --- as already pictured by
Leibniz --- be very fruitful for both areas.

The infrastructure that we have implemented for interactive and
automated reasoning in higher-order modal logics will clearly be
useful also outside philosophy; the range of potential applications is very wide.

\paragraph{Acknowledgements:} We thank Cedric Auger and Laurent Th\'ery, 
for their answers to our questions about Ltac in the Coq-Club mailing-list.

\bibliographystyle{plain}
%\bibliography{Bibliography}
\begin{thebibliography}{10}

\bibitem{Adams}
R.M.~Adams.
\newblock Introductory note to *1970.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1995.

\bibitem{AndersonGettings}
A.C.~Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

\bibitem{HOLND}
C.~Benzm{\"u}ller, C.~Brown, and M.~Kohlhase.
\newblock Higher-Order Semantics and Extensionality. 
\newblock {\em Journal of Symbolic Logic} 69(4):1027-1088, 2004.

\bibitem{JensOtten}
C.~Benzm{\"u}ller, J. Otten, and Th. Raths.
\newblock Implementing and Evaluating Provers for First-order Modal
Logics. 
\newblock In {\em ECAI 2012}, pages 163-168. IOS Press, 2012.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C.~Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications.

\bibitem{J23}
C.~Benzm{\"u}ller and L.C.~Paulson.
\newblock Quantified multimodal logics in simple type theory.
\newblock {\em Logica Universalis (Special Issue on Multimodal Logics)},
  7(1):7--20, 2013.

\bibitem{LEO-II}
C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke.
\newblock {LEO-II} - a cooperative automatic theorem prover for higher-order
  logic.
\newblock In {\em Proc. of IJCAR 2008}, volume 5195 of {\em LNAI}, pages
  162--170. Springer, 2008.

\bibitem{arXiv}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock Formalization, Mechanization and Automation of G{\"o}del's Proof of God's Existence.
\newblock {\em {arXiv}}:1308.4526, 2013.

\bibitem{AFP}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock G{\"o}del's God in Isabelle/HOL. 
\newblock {\em Archive of Formal Proofs}, 2013.

\bibitem{Coq}
Y.~Bertot and P.~Casteran.
\newblock {\em {Interactive Theorem Proving and Program Development}}.
\newblock Springer, 2004.

\bibitem{ModalLogic}
P.~Blackburn, J.v.~Benthem, and F.~Wolter (eds.), \newblock {\em {Handbook of Modal Logic}}. Elsevier, 2006.

\bibitem{Sledgehammer}
J.C.~Blanchette, S.~B\"ohme, and L.C.~Paulson.
\newblock Extending {Sledgehammer} with {SMT} solvers.
\newblock {\em Journal of Automated Reasoning}, 51(1):109--128, 2013.

\bibitem{Nitpick}
J.C.~Blanchette and T.~Nipkow.
\newblock Nitpick: A counterexample generator for higher-order logic based on a
  relational model finder.
\newblock In {\em Proc. of ITP 2010}, no.~6172 in LNCS, pages 131--146.
  Springer, 2010.

\bibitem{Satallax}
C.E.~Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
  Springer, 2012.

\bibitem{CIC}
Th.~Coquand and G.~Huet. 
\newblock The Calculus of Constructions. 
\newblock {\em Information and Computation} 76 (2-3). 1988.

\bibitem{ContemporaryBibliography}
R.~Corazzon.
\newblock Contemporary bibligraphy on the ontological proof
  (\url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}).

\bibitem{Fitting}
M.~Fitting.
\newblock {\em {Types, Tableaux and G\"odel's God}}.
\newblock Kluver Academic Press, 2002.

\bibitem{Labels}
D.M.~Gabbay. {\em Labelled Deductive Systems}. Clarendon Press, 1996.

\bibitem{Goedel1970}
K.~G\"odel.
\newblock Ontological proof.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1970.

\bibitem{GoedelNotes}
K.~G\"odel.
\newblock {\em Appendix A. Notes in Kurt G\"odel's Hand}, pages 144--145.
\newblock In  \cite{sobel2004logic}, 2004.

\bibitem{HOLLight}
J.~Harrison.
\newblock HOL Light: An overview.
\newblock In {\em Proceedings of TPHOLs 2009}. Number 5674 in LNCS, pp. 60-66. Springer, 2009.

\bibitem{Hazen}
A.P.~Hazen.
\newblock On G\"odel's ontological proof.
\newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

\bibitem{Hurd03first-orderproof}
J.~Hurd.
\newblock First-order proof tactics in higher-order logic theorem provers.
\newblock In {\em Design and Application of Strategies/Tactics in Higher Order
  Logics, NASA Tech. Rep. NASA/CP-2003-212448}, 2003.

\bibitem{Isabelle}
T.~Nipkow, L.C. Paulson, and M.~Wenzel.
\newblock {\em {Isabelle/HOL: A Proof Assistant for Higher-Order Logic}}.
\newblock Number 2283 in LNCS. Springer, 2002.

\bibitem{Zalta}
P.E.~Oppenheimera and E.N.~Zalta.
\newblock A Computationally-Discovered Simplification of the Ontological Argument.
\newblock {\em Australasian Journal of Philosophy}, 89(2):333-349, 2011.

\bibitem{Rushby}
J.~Rushby.
\newblock The Ontological Argument in PVS. 
\newblock In {\em Proc.~of CAV Workshop ``Fun With Formal Methods''}, St. Petersburg, Russia, 2013.

\bibitem{ScottNotes}
D.~Scott.
\newblock {\em Appendix B. Notes in Dana Scott's Hand}, pages 145--146.
\newblock In  \cite{sobel2004logic}, 2004.

\bibitem{sobel}
J.H.~Sobel. 
\newblock G\"odel's ontological proof. 
\newblock {\em In On Being and Saying. Essays for Richard Cartwright}, pages 241-261. MIT Press, 1987.

\bibitem{sobel2004logic}
J.H.~Sobel.
\newblock {\em Logic and Theism: Arguments for and Against Beliefs in God}.
\newblock Cambridge U. Press, 2004.

\bibitem{J22}
G.~Sutcliffe and C.~Benzm{\"u}ller.
\newblock Automated reasoning in higher-order logic using the {TPTP THF}
  infrastructure.
\newblock {\em Journal of Formalized Reasoning}, 3(1):1--27, 2010.

\bibitem{FormalTheologyRepository}
B.~Woltzenlogel Paleo and C.~Benzm\"uller.
\newblock Formal theology repository
  (\url{https://github.com/FormalTheology/GoedelGod}).
\end{thebibliography}

>>>>

\author{
  Christoph Benzm\"{u}ller\inst{1}\thanks{Supported by the German Research Foundation (DFG) under grant BE2501/9-1.} 
  \and 
  Bruno Woltzenlogel Paleo\inst{2}
}

@@

\begin{abstract}  
  This paper describes an embedding of higher-order modal logics in
  the \Coq proof assistant. {\Coq}'s capabilities are extended in a
  minimalistic manner, which is nevertheless sufficient for the
  formalization of significant, non-trivial modal logic proofs. The
  elegance, flexibility and convenience of this approach, from a user
  perspective, are illustrated here with the successful formalization
  of Gödel's ontological argument.
\end{abstract}

@@

Another recently explored possibility is the
embedding of first-order and even higher-order modal logics into classical
higher-order logics \cite{J23,B9,J22}, 
for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist. The embedding approach is flexible,
because various modal logics (even with multiple modalities or varying/cumulative domain quantifiers) can be easily supported by stating their
characteristic axioms. Moreover, the approach is relatively simple to
implement, because it does not require any modification in the source
code of the higher-order prover. The prover can be used as is, and
only the input files provided to the prover must be especially
encoded. Furthermore, the efficacy and efficiency of the embedding
approach has been confirmed in philosophically interesting and
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

However, one may wonder whether the embedding approach is adequate
also for \emph{interactive} reasoning, when the user proves theorems
by interacting with a proof assistant such as \Coq\footnote{The \Coq
proof assistant was chosen merely because of the authors' greater
familiarity with the tactic language of this system. Nevertheless, the
techniques presented and discussed here are likely to be useful for
other proof assistants, such as \texttt{Isabelle} \cite{Isabelle} or
\texttt{HOL-Light} \cite{HOLLight}, as well.} \cite{Coq}. The main goal and novelty
of this paper is to study this question. Our answer is positive.

@@

In order to illustrate the use of the implemented embedding, we show
here the formalization of Scott's version \cite{ScottNotes} of Gödel's
ontological argument for God's existence (in Section \ref{sec:proof}).
This proof was chosen mainly for two reasons. Firstly, it requires not
only modal operators, but also higher-order quantification. Therefore,
it is beyond the reach of specialized propositional and first-order
(modal) theorem provers. Secondly, this argument addresses an ancient
problem in Philosophy and Metaphysics, which has nevertheless received
a lot of attention in the last 15 years, because of the discovery of
the modal collapse \cite{Sobel}. This proof lies in the center of a
vast and largely unexplored application domain for automated and
interactive theorem provers.

The simpler ontological argument of Anselm has been automatically
verified with \texttt{PVS} by Rushby \cite{Rushby} and with 
first-order theorem provers by Oppenheimer and Zalta \cite{Zalta}. Gödel's
argument was automatically verified in our previous work on embedding-
based fully automated modal theorem proving \cite{arXiv,AFP}. This
paper presents the first fully interactive and detailed formalization
of this proof in a proof assistant. Hence we here reveal the proof
structure on a cognitively adequate level of
detail. Such detailed proof structure has been hidden in our
other papers on the subject due to the use of automated theorem
provers. In comparison to \cite{Rushby,Zalta}, however, our
contribution stands out with its surprising technical 
simplicity and elegance.

In addition to philosophy, propositional and quantified modal logics 
have (potential) applications in various other fields as well, including, for instance, 
verification, artificial intelligence
agent technologies, law and linguistics. \marginpar{BrunoToChris: Do we need citations for these applications?} 
Therefore, the main contribution described in this paper -- convenient techniques for leveraging a powerful proof assistant such as \Coq for interactive reasoning for modal logics -- may serve as a starting point for
many interesting projects.

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition mequal (x y: u)(w: i) := x = y.
\Notation "x m= y" := (mequal x y) (at level 99, right associativity).

\Definition mnot (p: o)(w: i) := ~ (p w).
\Notation "m~  p" := (mnot p) (at level 74, right associativity).

@@

\Definition mequiv (p q:o)(w:i) := (p w) <-> (q w).
\Notation "p m<-> q" := (mequiv p q) (at level 99, right associativity).
\end{Verbatim}

@@

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions (CIC) \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper. This calculus is classical, because of the double negation elimination rule. Although CIC is intuitionistic, it can be made classical by importing \Coq's classical library, which adds the axiom of the \emph{excluded middle} and the double negation elimination lemma.}
(e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

\begin{calculus}
{A (classical)  The natural deduction calculus}
{fig:NaturalDeduction}
\begin{small}

@@

\vspace{2em}

@@

\end{small}

@@

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they give rise to the inference
rules\footnote{The natural deduction calculus with the rules from
Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
complete relatively to the calculus of Figure
\ref{fig:NaturalDeduction} extended with a necessitation rule and the
modal axiom K \cite{ModalND}. Therefore,
assuming that the calculus from \ref{fig:NaturalDeduction} is sound
and Henkin-complete for classical higher-order logic \cite{NDHenkinCompleteness}, 
the additional modal rules in Figure \ref{fig:NDK} make it sound and 
Henkin-complete for modal higher-order logic \textbf{K}.} shown in Figure
\ref{fig:NDK}. The labels that name boxes in those inference rules are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\s\s\s\s
\infer[\nec_I]{\nec A}{\omega: \fbox{\infer*{A}{}} }
\s\s\s\s
\infer[\nec_E]{w: \fbox{ \infer*{}{A} } }{\nec A}

\vspace{2em}

\s\s\s\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s\s\s\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

@@

\noindent 
The proof of this lemma is displayed as a natural deduction
proof in  Figure \ref{fig:mp_dia}. As expected, \Coq's basic tactics
(e.g. \texttt{intros} and \texttt{apply}) work without modification.
The \texttt{intros p q H1 H2} tactic application corresponds to the
universal quantifier  and implication introduction inferences in the
bottom of the proof.  The \texttt{apply H3} tactic application
corresponds to the implication  elimination inference. The $\pos_E$,
$\pos_I$ and $\nec_E$ inferences correspond, respectively, to the
\texttt{dia\_e H1},  \texttt{dia\_i w0} and \texttt{box\_e H2 H3}
tactic applications.  The internal box named $w_0$ is accessed by
exactly one strong  modal inference, namely $\pos_E$.

\begin{figure}[H]
\centering
\label{fig:mp_dia}
\includegraphics[width=0.5\textwidth]{ND-(mp_dia).png}
\caption{Natural deduction proof of \texttt{mp\_dia}}
\end{figure}

\noindent
The same lemma could be proved without the new modal tactics, as shown below.
But this is clearly disadvantageous, for several reasons: 
the proof script becomes longer;
the definitions of modal operators must be unfolded, 
either explicitly (as done below) or implicitly in the user's mind; 
tactic applications dealing with modal operators cannot be easily distinguished
from tactic applications dealing with quantifiers; and  
hypotheses about the reachability of worlds (e.g. \texttt{R1} below) 
must be handled explicitly. In summary, without the modal tactics, 
a convenient and intuitive correspondence between proof scripts and 
modal natural deduction proofs would be missing.

@@

In order to demonstrate the efficacy and convenience of the modal
embedding approach not only for proving simple lemmas and theorems,
but also for larger developments, we include here a full and detailed
formalization of Gödel's ontological argument.

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology. G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern
culmination of this tradition, following particularly the footsteps of
Leibniz. Various slightly different versions of axioms and definitions
have been considered by G\"{o}del and by several philosophers who
commented on his proof (cf.
\cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).
The formalization shown in this Section aims at being as similar as
possible to Dana Scott's version of the proof \cite{ScottNotes}. The
formulation and numbering of axioms, definitions and theorems is the
same as in Scott's notes. Even the \Coq proof scripts follow precisely
all the steps in Scott's notes. Scott's assertions are emphasized
below with comments. In contrast to the formalization in \Isabelle
\cite{AFP}, where automation via \texttt{Metis}
\cite{Hurd03first-orderproof} and \texttt{Sledgehammer}
\cite{Sledgehammer} using tools such as \texttt{Nitpick}
\cite{Nitpick}, \texttt{LEO-II} \cite{LEO-II} and \texttt{Satallax}
\cite{Satallax} has been successfully employed, the formalization in
\Coq used no automation. This was a deliberate choice, mainly because
it allowed a qualitative evaluation of the convenience of the
embedding approach for \emph{interactive} theorem proving. Moreover,
in order to formalize exactly Scott's version and not some arbitrary 
version found automatically, automation would have to be heavily
limited anyway. Furthermore, the deliberate preference for simple
tactics (mostly \emph{intro}, \emph{apply} and the modal tactics
described in Section \ref{sec:Tactics}) results in proof scripts that
closely correspond to common natural deduction proofs. This hopefully makes
the formalization more accessible to those who are not experts in
\Coq's tactics but are nevertheless interested in G\"odel's proof.

Gödel's proof requires \Coq's classical logic libraries as well as the
\texttt{Modal} library developed by us and described in Sections
\ref{sec:Embedding} and \ref{sec:Tactics}.

@@

\noindent
Gödel's theory has a single higher-order constant, \texttt{Positive}, 
which ought to hold for properties considered \emph{positive} in a moral sense.

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\com{(* Axiom A1:} 
\com{   either a property or its negation is positive, but not both *)}
\Axiom axiom1a : 
  [ mforall p, (Positive (\fun x: u => m~(p x))) m-> (m~ (Positive p)) ].

@@

Although fully automated (as opposed to interactive) theorem proving
is beyond the scope of this paper, it is worth mentioning that all
lemmas and theorems in Sections \ref{sec:Embedding},
\ref{sec:SimpleExamples} and \ref{sec:BeyondK} could have been proven
automatically using \Coq's \texttt{firstorder} tactic. The
implementation of hints to allow \Coq's automatic tactics to take full
advantage of the embedding and the modal axioms still remains for
future work.

The theological implications of the verified correctness of Gödel's
proof certainly depend on a critical discussion of the underlying
concepts, definitions and axioms, which is beyond the scope of this
paper.  Clearly, the application of theorem proving technology to the
domain of theoretical philosophy can --- as already pictured by
Leibniz --- be very fruitful for both areas.

The infrastructure that we have implemented for interactive and
automated reasoning in higher-order modal logics will clearly be
useful also outside philosophy; the range of potential applications is
in fact very wide.

\paragraph{Acknowledgements:} We thank Cedric Auger and Laurent Théry, 
for their answers to our questions about Ltac in the Coq-Club mailing-list.

\bibliographystyle{plain}
\bibliography{Bibliography}
